---
title: "Afternoon Exercises"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)

# read these files from the data folder
# but don't include in output 

# Read in PM_Borders1.csv

PM_Borders1 <- read_csv("../data/PM_Borders1.csv")

# Read in PM_Borders2.csv

PM_Borders2 <- read_csv("../data/PM_Borders2.csv")
```

1. Read in both PM_Borders1.csv and PM_Borders2.csv separately using `read_csv()`. Remember to load the `readr` package first.
<br>

```{r, warning=FALSE, message=FALSE, eval = FALSE}
library(readr)

# Read in PM_Borders1.csv

PM_Borders1 <- read_csv("data/PM_Borders1.csv")

# Read in PM_Borders2.csv

PM_Borders2 <- read_csv("data/PM_Borders2.csv")
```

<br>

2. `PM_Borders1` and `PM_Borders2` have the same columns and there are some records that exist within both files. Choose an appropriate join function to merge these data frames together and call this `joined_data`. 

<br>

```{r, warning=FALSE, message=FALSE}
# Use full_join to merge these files

joined_data <- full_join(PM_Borders1, PM_Borders2)

# Show data

joined_data

# Compare dimensions

dim(PM_Borders1)
dim(PM_Borders2)
dim(joined_data)
```

<br>

3. Using `joined_data`, create a crosstab of `ManagementofPatient` and `Specialty`.

<br>

```{r}
# Use table() to get the crosstab

table(joined_data$ManagementofPatient, joined_data$Specialty)
```

<br>

4. Take all records in `joined_data` which have `ManagementofPatient` values "Day Case in Inpatient Ward" or "Day Case in Day Bed Unit" and store them in a dataframe called `day_cases`. Calculate a new column `LOS2` for `LengthOfStay` multiplied by two.

<br>

```{r }
# Use filter() to select values and mutate() to calculate LOS2

day_cases <- joined_data %>% 
  filter(ManagementofPatient == "Day Case in Inpatient Ward" | ManagementofPatient == "Day Case in Day Bed Unit") %>% 
  mutate(LOS2 = LengthOfStay * 2)

# Show data

day_cases %>% select(ManagementofPatient, LengthOfStay, LOS2)
```

<br>

5. Group `joined_data` by `Specialty` and then calculate the mean `LengthOfStay` value for each `Specialty`. Arrange the data by mean `LengthOfStay` from largest to smallest. 

<br>

```{r}
# Use group_by() to group the data and then use summarise() to calculate the mean length of stay. Use arrange() to sort the data

joined_data %>% 
  group_by(Specialty) %>% 
  summarise(mean_los = mean(LengthOfStay)) %>% 
  arrange(desc(mean_los))
```

<br>

6. Select the `LinkNo`, `HospitalCode` and `Main_op` columns in `joined_data` and rename the `Main_op` column to `Main_Operation`. Store this data in a dataframe called `filtered_data` and only keep records that do not have a missing value for `Main_Operation`. How many records have been removed?

<br>

```{r }
# Use select() to select columns and rename() to rename Main_op
# Use filter() and !is.na() to obtain records which do not have a missing value

filtered_data <- joined_data %>% 
  select(LinkNo, HospitalCode, Main_op) %>% 
  rename(Main_Operation = Main_op) %>% 
  filter(!is.na(Main_Operation))

nrow(joined_data) - nrow(filtered_data)
```

<br>

7. Plot a histogram of `LengthOfStay` for records in `joined_data` with a `DateofAdmission` in April 2015. Give your histogram a suitable title, colour and axis labels.

<br>

```{r}
hist_data <- joined_data %>%
  filter(DateofAdmission >= 20150401 & DateofAdmission <= 20150430)

hist(hist_data$LengthOfStay, col = "lightblue", main = "Histogram of Length of Stay for April 2015", xlab = "Length of Stay", ylab = "Frequency")
```

